### 2025-09-25 paper/sec/introduction.tex

When attempting to interpret or verify data-driven claims, a key challenge lies in tracing specific claims
back to the relevant data. In peer review, for example, empirical claims typically lack author-supplied links
to data, making them hard for reviewers to check directly~\citep{weber20}. Paper retractions, meanwhile, are
often attributable not to fraud, but to simple errors in data management or analysis~\citep{hu25}. The use of
large language models (LLMs) to interpret scholarly documents has seen considerable attention recently, from
fact-checking~\citep{abu-ahmad25} to interpretation of charts and figures~\citep{roberts24}, but current LLM
interfaces do not support direct interrogation of visual or other outputs for traceability to inputs.

@@

This transparent version of the document was implemented in Fluid. The source code is shown in
\figref{fluid-example-paragraph}, and makes use of several helper functions, a representative subset of which
are shown in \figref{fluid-scigen}. Both the provenance-tracking runtime of Fluid \emph{and} the LLM-based
authoring support contribute to the solution: the Fluid runtime provides the interactions, and the LLM-based
authoring tool makes the whole process practical. Without tool support, authoring a scientific document as a
(hand-crafted) \emph{program} is unlikely to be a feasible task.

>>>>

When interpreting or verifying data-driven claims, a key challenge lies in tracing specific claims back to the
relevant data. In peer review, for example, empirical claims typically lack author-supplied links to data,
making them hard for reviewers to check directly~\citep{weber20}. Paper retractions, meanwhile, are often
attributable not to fraud, but to simple errors in data management or analysis~\citep{hu25}. The use of large
language models (LLMs) to interpret scholarly documents has seen considerable attention recently, from
fact-checking~\citep{abu-ahmad25} to interpretation of charts and figures~\citep{roberts24}, but current LLM
interfaces do not support direct interrogation of visual or other outputs for traceability to inputs.

@@

This transparent version of the document was implemented in Fluid. The source code is shown in
\figref{fluid-example-paragraph}, and makes use of several helper functions, a representative subset of which
are shown in \figref{fluid-scigen}. What makes our solution interesting is that the provenance-tracking
runtime of Fluid \emph{and} the LLM-based authoring support are both essential components of the solution,
with Fluid providing the interactions, and the LLM-based tool making the authoring process feasible.
Generating code for a traditional language like Python would still result in a data-driven document, but
crucially without the interactive provenance queries; and without AI-based tooling to support the authoring
process, the author would be faced with creating the code in \figref{fluid-example-paragraph} by hand, which
is unlikely to be feasible as part of the usual scientific writing process.
